A seamless, client-centric programming model for type safe web applications

### 2014-05-08 paper.tex
Restructure implementation stuff

To run this example, an address and a port must be provided so that the client
knows which server to contact. There are several ways of doing this:
using the GHC plugin system, through Template Haskell or by slightly altering
how program entry points are treated in a compiler or wrapper script, to name
a few. A non-intrusive method when using the GHC/Haste compiler pair would
be to pass add \lstinline!-main-is setup! to both compilers' command line
and add the \lstinline!setup! function in figure \ref{lst:setup} to the source
code. This will instruct the server binary to listen on the port 1111 when
started, and the client to attempt contact with that port on the local machine.

@@

\begin{listingfloat}
\begin{code}
setup :: IO ()
setup =
  runApp (mkConfig "ws://localhost:1111" 1111) main
\end{code}
\caption{Bootstrapping a Haste.App program}
\label{lst:setup}
\end{listingfloat}

@@

Aside from \lstinline!runClient!, the \lstinline!App! monad has two primary
operations, both of which have slightly different implementations on the client
and server respectively.

\begin{code}
data RemoteValue a = Here a | There
data Remote a = Remote CallID [JSON]

liftServerIO :: IO a -> App (RemoteValue a)
remote :: Remotable a => a -> App (Remote a)
\end{code}

The client side definition of \lstinline!liftServerIO! is as simple as can be.

\begin{code}
liftServerIO _ = App $ return There
\end{code}

Its argument is simply ignored, and an opaque dummy value is returned; recall
that server-lifted IO computations are not observable on the client. The
server side implementation is only marginally more complex.

\begin{code}
liftServerIO m = App $ fmap Here m
\end{code}

Its argument IO computation is executed, and the result is wrapped in the
\lstinline!Remote! constructor. The counterpart function, \lstinline!use!
simply unwraps the \lstinline!Here! constructor and returns the contained
value in the \lstinline!Server! monad.

\begin{code}
use :: Remote a -> Server a
use (Here x) = return x
\end{code}

\lstinline!remote! is a more interesting function. Looking at its type, we see
that it accepts any \lstinline!Remotable! value for its argument.
\lstinline!Remotable! is defined as follows.

@@

The client implementation of \lstinline!remote! is not very complex.

\begin{code}
remote _ = App $ do
  (next_id, remotes) <- get
  put (next_id+1, remotes)
  return (Remote next_id [])
\end{code}

As the remote function lives on the server, the client only needs an
identifier to be able to call on it. The remote function is thus ignored,
so that it can be optimized out of existence in the client version of the code.
The server version is only marginally more complex in that it updates the
identifier-function mapping table as well.

@@

\begin{code}
remote f = App $ do
  (next_id, remotes) <- get
  put (next_id+1, (next_id, mkRemote f) : remotes)
  return (Remote next_id [])
\end{code}

The second argument of the \lstinline!Remote! constructor is the list of
arguments to be sent to the function, to be updated as the remote function is
applied using the \lstinline!<.>! operator.

After the \lstinline!App! computation finishes, the identifier-function mapping
accumulated in its state is handed over to the server's event loop, where it is
used to dispatch the proper functions for incoming calls from the client.

>>>>

To run this example, an address and a port must be provided so that the client
knows which server to contact. There are several ways of doing this:
using the GHC plugin system, through Template Haskell or by slightly altering
how program entry points are treated in a compiler or wrapper script, to name
a few. A non-intrusive method when using the GHC/Haste compiler pair would
be to pass add \lstinline!-main-is setup! to both compilers' command line
and add the \lstinline!setup! function to the source code.
This will instruct the server binary to listen on the port 1111 when
started, and the client to attempt contact with that port on the local machine.

\begin{code}
setup :: IO ()
setup =
  runApp (mkConfig "ws://localhost:1111" 1111) main
\end{code}

@@

@@

The workings of the \lstinline!App! monad basically hinges on two abstract data
types: \lstinline!RemoteValue! and \lstinline!Remote!. \lstinline!RemoteValue!
is used to represent values which are available to the server but not to the
client. \lstinline!Remote! denotes functions which live on the server but can
be invoked remotely by the client. The implementation of these types and the
functions that operate on them differ between the client and the server.

\paragraph{Client side implementations}
We begin by looking at the client side implementation.

\begin{code}
data RemoteValue a = There
data Remote a = Remote CallID [JSON]
\end{code}

As values of the \lstinline!RemoteValue! type are not accessible to the client,
they are represented by a simple dummy value. The \lstinline!Remote! type is
slightly more useful, containing the ID of a remote function and a list of the
serialized arguments to be passed when invoking it. In essence, the
\lstinline!Remote! type is an explicit representation of a remote closure.
Such closures can be applied to values using the \lstinline!<.>! operator.

The \lstinline!remote! function is used to bring server side functions into
scope on the client as \lstinline!Remote! functions. It is implemented using a
simple counter which keeps track of how many functions have been imported so
far and thus which ID to assign to the next remote function.

\begin{code}
remote :: Remotable a => a -> App (Remote a)
remote _ = App $ do
  (next_id, remotes) <- get
  put (next_id+1, remotes)
  return (Remote next_id [])
\end{code}

As the remote function lives on the server, the client only needs an
identifier to be able to call on it. The remote function is thus ignored,
so that it can be optimized out of existence in the client executable. Looking
at its type, we can see that \lstinline!remote! accepts any argument
instantiating the \lstinline!Remotable! class.
\lstinline!Remotable! is defined as follows.

@@

The final function operating on these types is \lstinline!liftServerIO!, used
to initialize state holding elements and perform other setup functionality on
the server.

\begin{code}
liftServerIO :: IO a -> App (RemoteValue a)
liftServerIO _ = App $ return There
\end{code}

As we can see, the implementation is as simple as can be. Since
\lstinline!RemoteValue!s are represented by a dummy value on the client, we
just return said value.

\paragraph{Server side implementations}
The server side representation of the \lstinline!RemoteValue! and
\lstinline!Remote! types are in a sense the opposites of their client side
counterparts.

\begin{code}
data RemoteValue a = Here a
data Remote a = Remote
\end{code}

Where the client is able to do something useful with the \lstinline!Remote!
type but not with \lstinline!RemoteValue! the server has no way to inspect
\lstinline!Remote! functions, and thus only has a no-op implementation of the
\lstinline!<.>! operator. On the other hand, the server has full access to the
contents of \lstinline!RemoteValue!s via the \lstinline!use! function.

\begin{code}
use :: RemoteValue a -> Server a
use (Here x) = return x
\end{code}

Those values are produced by the \lstinline!remoteIO! function, which is quite
simple. It merely executes its argument and wraps its argument in the
\lstinline!RemoteValue! type.

\begin{code}
remoteIO :: IO a -> Server (RemoteValue a)
remoteIO m = App $ fmap Here m
\end{code}

The server version of \lstinline!remote! is a little more complex than its
client side counterpart. In addition to keeping track of the identifier of the
next remote function, the server side \lstinline!remote! pairs up remote
functions with these identifiers in an identifier-function mapping.

\begin{code}
remote f = App $ do
  (next_id, remotes) <- get
  put (next_id+1, (next_id, mkRemote f) : remotes)
  return Remote
\end{code}

@@

\paragraph{The server side dispatcher}
After the \lstinline!App! computation finishes, the identifier-function mapping
accumulated in its state is handed over to the server's event loop, where it is
used to dispatch the proper functions for incoming calls from the client.
